Multi-level music structure analysis: an audio-based approach using deep learning

### 2023-02-05 introduction/introduction.tex
Updates from Overleaf

Music Structure Analysis (MSA) is a field of study that aims to understand music's underlying organization and structure. This can include analyzing elements such as melody, harmony, rhythm, and timbre, as well as the relationships and interactions between these elements. MSA aims to identify patterns and regularities in music and develop computational models that can replicate and generate musical structures.

The fundamental concept behind MSA is the ability to divide a song into distinct, non-overlapping segments, assigning a label to each segment that defines its type. This segmentation and labeling process aims to capture a human's perception or analysis of the song.

MSA is a multidisciplinary field that draws on techniques from music theory, signal processing, machine learning, and other fields. It can be applied to various musical styles and genres and has been used for tasks such as music transcription, generation, and similarity.

\section{Historical overview}

Music Structure Analysis (MSA) has a long history, with its roots dating back to music theory and composition studies. In the 18th and 19th centuries, music theorists such as Rameau~\cite{christensen2004rameau} and Schenker~\cite{schenkerdocumentsonline} developed analytical frameworks for understanding the structure of music in terms of harmony, melody, and form.

@@

@@

As a music lover and connoisseur, I am interested in gaining a deeper understanding of the fundamental building blocks of musical composition and the various techniques used to create musical works~\cite{choi2017convolutional}. I am deeply interested in learning how to retrieve embedded and encoded information within music. The study of music structure analysis provides a means to systematically analyze and understand musical pieces. It has traditionally been approached through music theory, formal musical notation, and musical analysis techniques.

However, with the advent of artificial intelligence and machine learning, the field of music structure analysis has expanded to encompass the use of musically motivated neural network architectures. These networks are designed to incorporate musical concepts and perceptual approaches and use this knowledge to analyze and understand the structure of musical pieces in a way that goes beyond traditional music analysis techniques.

By studying music structure analysis using musically motivated neural networks, you will gain an in-depth understanding of the latest advancements in the field and how these techniques are being used to revolutionize our understanding of music. Additionally, you will be able to contribute to developing new and innovative methods for musical analysis~\cite{Huang2019MusicTG}

>>>>

Music Structure Analysis (MSA) is the field of study that aims to understand music's underlying organization and structure. This can include analyzing elements such as melody, harmony, rhythm, and timbre, as well as the relationships and interactions between these elements. MSA aims to identify patterns and regularities in music and develop computational models that can replicate and generate musical structures.

The fundamental concept behind MSA is the ability to divide a song into distinct, non-overlapping segments, assigning a label to each component that defines its type. This segmentation and labeling process aims to capture a human's perception or analysis of the song.

MSA is a multidisciplinary field that draws on techniques from music theory, signal processing, machine learning, and other areas. It can be applied to various musical styles and genres and has been used for tasks such as music transcription, generation, and similarity.

\section{Topic overview}

Music Structure Analysis (MSA) has a long history, dating back to music theory and composition studies. In the 18th and 19th centuries, music theorists such as Rameau~\cite{christensen2004rameau} and Schenker~\cite{schenkerdocumentsonline} developed analytical frameworks for understanding the structure of music in terms of harmony, melody, and form.

@@

As a music lover and connoisseur, I am interested in gaining a deeper understanding of the fundamental building of musical composition and the various techniques used to create musical works~\cite{choi2017convolutional}. I am deeply interested in learning how to retrieve embedded and encoded information within the music. The study of music structure analysis provides a means to systematically analyze and understand musical pieces. It has traditionally been approached through music theory, formal musical notation, musical analysis techniques, and machine learning, which I plan to blend.

With the advent of artificial intelligence and machine learning, the field of music structure analysis has expanded to encompass the use of musically motivated neural network architectures. These networks are designed to incorporate musical concepts and perceptual approaches and use this knowledge to analyze and understand the structure of musical pieces in a way that goes beyond traditional music analysis techniques.

As a professional music transcriber, you are interested in understanding how deep neural networks (DNNs) filter musical features. In other words, I am eager to learn how DNNs listen~\cite{7500246} to music.

By studying music structure analysis using musically motivated neural networks, you will gain an in-depth understanding of the latest advancements in the field and how these techniques are being used to revolutionize our understanding of music. Additionally, we will be able to contribute to developing new and innovative methods for musical analysis~\cite{Huang2019MusicTG}

@@

As a professional music transcriber, I aim to understand how deep neural networks (DNNs) filter musical features, listen to the NN layers from a musician's perspective and analyze the resulting data. My goal is to approach the technology from both a technical and musical standpoint to understand how DNNs process and extract musical information. By doing so, you aim to enhance the accuracy and efficiency of the process and contribute to developing AI technology for music analysis. By combining my expertise in music with my -not fully developed- knowledge of DNNs, I hope to bring a unique perspective to the field and contribute to the advancement of music analysis.

### 2023-02-05 music analysis/authors.tex
Updates from Overleaf

What we're trying for is a very high overview of musical development in terms of a vocabulary constantly being enriched by more and more remote and chromatic overtones. It's as if we could see the whole of music developing from prehistory to the present in two minutes. Let's again pretend we're hominids, and that the smash hit of the moment is, let's say, Fear Harvard. Here we are in our hominid hut, crooning, Fear Harvard relations to age, and now maybe our wives and maybe our prepubescent sons join in, and automatically we're singing not in unison, but in octaves, since men's and women's voices are naturally an octave apart. Now that octave interval, I wish I could sing an octave so I could really show you what I mean, but that octave interval happens to be the first interval of the harmonic series as you remember, right? Okay, now centuries pass, and the next interval of the harmonic series is assimilated by humanity. Namely, the fifth. And now we can be singing this, by those festival rides on the horizon. Now, of course, this little change brings us forward a mere 10 million years into the 10th century AD and into a fairly sophisticated musical culture. But now we admit the next interval of the series, the fourth, and now we can mix intervals of the octave and the fifth and the fourth. That's beginning to sound like polyphony. And again comes a great leap as music absorbs the next overtone, the third, and just listen to the difference. It's a whole new music, richer, mellower, with a new coloristic warmth, and as we know, this new interval of the third, because I like the older sound better, but anyway, as we know, this new interval of the third introduces into music the phenomenon of the triad so that now, fair Harvard, can begin to sound more like its Victorian self. And so there is born what we now call tonal music, a stable tonal language firmly rooted in the basic notes of the harmonic series, the fundamental and its first different overtone, the fifth, now and forever more to be known as the tonic and the dominant. And that fifth interval really does dominate, because once this tonic-dominant relationship is established, it's a field day for composers. There can now be fifths of fifths of fifths of fifths, each one of them a new tonic producing a new dominant, a whole circle of fifths, 12 of them in fact, always winding up with the starting tone, where the preceding upwards, let's say from the low C, G, D, A, E, B, F, sharp, C sharp, G sharp, D sharp, B flat, F, C, that's again C, believe it or not. Or preceding downwards, starting from that, C, F, B flat, E flat, E flat, G flat, G flat, C, back to C. That's a circle of 12 fifths, and that's the answer I promised you. That's how we get the 12 different tones of our chromatic scale. In other words, if you take all those 12 of the circle of fifths and put them together in scale order, you'll get this. And what's more, those 12 notes generate a circle of 12 keys, through which, thanks to the perfecting of the temperate system, composers can now go freewheeling at their own chromatic pleasure. Now this means that ultimately, fair Harvard can sound like this. Now that's chromatic porridge, and in our own century it's going to become goulash. How does music contain this loose, runny chromaticism? By the basic principle of diatonicism, that stable relationship of tonics, and dominant, subdominants, and supertonics, and new dominants, and new tonics. We can now modulate as freely as we want, as chromatically as we want, and still have complete tonal control. This great system of tonal controls was perfected and codified by Bach, Johann Sebastian Bach, whose genius was to balance so delicately and so justly these two forces of chromaticism and diatonicism, forces that were equally powerful and presumably contradictory in nature.

>>>>

\chapter{Music Structure Analysis (MSA)}

\section{The music structure analysis problem}
Audio-based Music Structural Analysis (MSA) identifies adjacent and non-overlapping musical segments in an audio signal and labels them based on similarity. Segments can occur at various time scales, from short motives to longer sections. Flat MSA focuses on non-hierarchical structures, while hierarchical MSA considers multiple time scales. However, MSA faces challenges such as boundary placement ambiguity and similarity quantification. Segments must be non-overlapping for a given hierarchical level, although some musicological approaches allow for overlap.

\subsection{Problem definition}

The musical structure analysis (MSA) challenges the definition due to subjectivity, ambiguity, and data scarcity. A flat structural analysis of an audio signal consists of a set of contiguous, non-overlapping, and labeled time intervals that span the entire duration of the audio signal. Given a musical recording of T audio samples, a flat segmentation is defined by a set of segment boundaries B, a set of k unique labels Y, and a mapping of segment starting points to labels S. This leads to the label assignment L, which assigns a label to each time point. The segment boundaries are usually sampled at a rate of 10 Hz to achieve a balance between resolution and computational efficiency.

As an art form, music is open to interpretation and can evoke different emotions and meanings in different listeners. This subjectivity makes it difficult to objectively analyze the structure of a piece of music, as different annotators may place boundaries between segments differently based on their perception and interpretation of the music. It can also have multiple structures, including hierarchical structures that make it difficult to untangle the truth.

\subsection{Segmentation principles}
The principles of musical structure analysis (MSA) were initially defined as homogeneity, novelty, and repetition, with the addition of regularity later on. To analyze the structure of a musical recording, a self-similarity matrix (SSM) is often used to visualize the degree of similarity between audio frames. The SSM results in a square matrix with the highest degree of similarity on the diagonal.

\subsubsection{Homogeneity and novelty}
Homogeneity assumes that segments are homogeneous in musical attributes, with boundaries between dissimilar segments being points of novelty.

\subsubsection{Repetition}
Repetition assumes that segments with the same label are similar sequences, with boundaries defined by repeated sequences' start and endpoints. The structure of the music is represented using a self-similarity matrix to make its structure visually apparent. Homogeneous blocks may be challenging to subdivide, while repeated sequences appear as paths in the self-similarity matrix. MSA approaches focus on identifying these repeated sequences.

\subsubsection{Regularity}
The regularity principle in music structure analysis aims to exploit the regularities present in musical segments, such as the log-normally distributed length across the track and the tendency for the duration of two equally labeled segments to span an integer ratio of beats.

\section{Theoretical approach}
\subsection{Heinrich Schenker}

Heinrich Schenker~\cite{schenkerdocumentsonline} was a prominent music theorist who developed \textit{Schenkerian Analysis}, a method for analyzing tonal music. This method is based on the concept of "the Ursatz," a fundamental structure of three levels: foreground, middle-ground, and background. Schenker's approach emphasizes the horizontal dimension of music, specifically melody, and the relationships between individual notes and the piece's overall structure. His ideas have significantly impacted the field of music theory and analysis, and the method is still widely used today. While the approach is primarily theoretical and relies on human interpretation, it can also be approached through computer algorithms such as CNNs; the results are not the same as human interpretation.

\subsection{Schenkerian analysis}

Schenkerian analysis is a musical analysis method that examines tonal music using the theories of Heinrich Schenker. The purpose is to showcase the organic unity of a piece by highlighting the relationship between the foreground (all musical notes in the score) and the deep abstract structure known as the \textit{Ursatz}. The Ursatz, roughly the same for all tonal music, is then developed into a unique foreground in each work. The concept of tonal space, defined as the intervals between the notes of the tonic triad in the background, is crucial to the analysis. The tonal space is filled with passing and neighbor tones, creating new triads and tonal spaces for further elaboration.

Schenkerian analysis uses specialized musical notation and is either presented in a generative manner, starting from the Ursatz and ending at the score or in a reductive manner, starting from the score and reducing it to its fundamental structure. The Ursatz is presented in an arrhythmic graph, and rhythmic signs display hierarchical relationships between pitch events.

While complex and challenging, Schenkerian analysis is designed to reveal the internal coherence of a work, which ultimately lies in its tonality. The analysis's results may reflect the analyst's perceptions and intuitions.

### 2023-02-05 music analysis/music structure analysis.tex
Updates from Overleaf

In Schenkerian analysis, the fundamental structure, also known as Ursatz, represents the underlying structure of a tonal work at the most remote or background level, in the most abstract form. It comprises an essential elaboration of the tonic triad, including the fundamental line and accompanying bass arpeggiation. The fundamental structure, like the fundamental line, takes one of three forms, depending on which pitch of the tonic triad serves as the primary tone. For example, in the key of C major, the fundamental structure would be represented by the descending fundamental line from scale degree 3.

@@

The upper voice of the fundamental structure, represented by the fundamental line, utilizes a descending direction. In contrast, the lower voice, the bass arpeggiation through the fifth, takes an ascending direction (as shown in figure 1). The combination of the fundamental line and bass arpeggiation constitutes a unity, and neither can exist alone; it is only through their unification in a contrapuntal structure that they produce art.

@@

In Schenkerian analysis, structural levels refer to the representation of a piece of music at different levels of abstraction. These levels typically include the foreground, middle ground, and background. Schenker posits that musical form, as represented through these structural levels, can be understood as an energy transformation, specifically as the transformation of forces that flow from the background to the foreground through the various levels.

@@

Music form refers to the structural organization of a musical composition or performance. This can include elements such as the arrangement of musical units (such as rhythm, melody, and harmony) that exhibit repetition or variation, the arrangement of instruments, or the orchestration of a symphonic piece. These elements interact to shape the composition and create a meaningful musical experience for the listener. The organizational elements can be further broken down into smaller units called phrases, which express a musical idea but lack the sufficient weight to stand alone. The form of a composition unfolds over time through the expansion and development of these ideas. In tonal harmony, form is primarily articulated through cadences, phrases, and periods.

>>>>

In Schenkerian analysis, the fundamental structure, also known as Ursatz, represents the underlying structure of a tonal work at the most remote or background level, in the most abstract form. It comprises an essential elaboration of the tonic triad, including the entire line and accompanying bass arpeggiation. The fundamental structure, like the fundamental line, takes one of three forms, depending on which pitch of the tonic triad serves as the primary tone. For example, in the key of C major, the fundamental structure would be represented by the fundamental descending line from scale degree 3.

@@

The upper voice of the fundamental structure, represented by the whole line, utilizes a descending direction. In contrast, the lower voice, the bass arpeggiation through the fifth, takes an ascending direction (as shown in figure 1). The combination of the fundamental line and bass arpeggiation constitutes a unity, and neither can exist alone; it is only through their unification in a contrapuntal structure that they produce art.

@@

In Schenkerian analysis, structural levels refer to the representation of a piece of music at different levels of abstraction. These levels typically include the foreground, middle ground, and background. Schenker posits that musical form, as represented through these structural levels, can be understood as an energy transformation, specifically the transformation of forces that flow from the background to the foreground through the various levels.

@@

Music form refers to the structural organization of a musical composition or performance. This can include elements such as the arrangement of musical units (such as rhythm, melody, and harmony) that exhibit repetition or variation, the arrangement of instruments, or the orchestration of a symphonic piece. These elements interact to shape the composition and create a meaningful musical experience for the listener. The organizational elements can be further broken down into smaller units called phrases, which express a musical idea but lack sufficient weight to stand alone. The form of a composition unfolds over time through the expansion and development of these ideas. In tonal harmony, form is primarily articulated through cadences, phrases, and periods.
